Making Soup: Preparing and Validating Molecular Simulations of the Bacterial Cytoplasm

### 2019-10-17 paper/virtual_cytoplasm_v2.tex
Updated for submission

\newpage
\begin{abstract}
Biomolecular crowding affects the biophysical and biochemical behavior
of macromolecules when compared to the dilute environment present in
experiments made with isolated proteins. Computational modeling and
simulation are useful tools to study how crowding affects the
structural dynamics and biological properties of macromolecules. As
computational power increased, modeling and simulating large scale
all-atom explicit solvent models of the prokaryote cytoplasm become
possible. In this work, we build an atomistic model of the cytoplasm
of \textit{Escherichia coli} composed of 1.5 million atoms and submit
it to a total of \SI{3}{\micro\second} of molecular dynamics
simulations. The simulation model is consistent with experimental data
about the diffusion coefficient and stability of macromolecules under
crowded conditions. In order to stimulate further work we provide a
Python script and a set of files that enables any researcher to build
their own \textit{E. coli} cytoplasm models to address questions
related to crowding.
\end{abstract}

@@

The intracellular environment is markedly different from the
\textit{in vitro} model systems from which most of our knowledge about
the biochemical and biophysical behavior of macromolecules has been
derived~\cite{Feig2017a}. The \textit{in vivo} microenvironment is
heterogeneous with high concentrations of many different compounds,
leading to spatial constrains that directly affects biological
activities~\citep{ostrowska2019}. This effect, called macromolecular
crowding, influences the stability of many macromolecules
because the equilibrium is shifted towards a stable native structure
as a result of the entropic component given by the excluded volume
effect~\cite{cheung2005}. At the same time, an effective
destabilization due to the enthalpic component of the Gibbs energy has
been observed, due to replacing interactions with water by weaker
interactions with the
crowders~\cite{Feig2011,miklos2011,Wang2012b}. Additionally, crowding
affects the structure and dynamics of water around proteins, which has
a significant effect on their biological
activities~\cite{Harada2012a,king2013}.

While the structure and dynamics of biomolecules is well characterized
\textit{in vitro}, the understanding of the influences of crowded
environments \textit{in vivo} are still evolving.  Techniques such as
nuclear magnetic resonance~\cite{reckel2007,pielak2008} and
fluorescence spectroscopy~\cite{ignatova2004,xie2008,English2011} are
among the most relevant to the field due to their ability to probe
cells in a non destructive way. Another essential class of methods
come from computational modeling and simulation. By extrapolating the
advances in computational power in the last decades to the future, it
was estimated that the atomistic simulation of an entire
\textit{Escherichia coli} bacterium for \SI{1}{\nano\second} will be
possible in 15 years from now~\cite{vanGunsteren2006a}.
% {\bf THAT WAS 13 YEARS AGO. WHAT IS THE STATE OF THE ART?} {\color{blue} The paper predicts that simulating a whole E. coli bacterium with $10^{11}$ atoms for 1 ns will be possible in 2034, thus the "15 years from now" in our text. That was meant as an introduction to the simulation of big systems. The evolution of computational models to study crowing and the state-of-the-art is discussed in the next paragraph. Maybe we can move this phrase to the beginning of the next paragraph }

As the available computational power increases, models become
increasingly complex and are better representations of reality. The
most simple class of computational models apply Brownian Dynamics
calculations to probe macromolecules immersed in a solution of
crowders that are treated as hard spheres with varying radii to
represent different types of molecules~\cite{Ando2010}. These models
usually use an implicit representation of the
solvent~\cite{Mcguffee2010}. More recently, all-atom explicit solvent
molecular dynamics simulations have been employed on crowded systems
composed of several copies of one or a few
macromolecules~\cite{Wang2017c}. However, these models do not reflect
the true heterogeneity of biologically relevant crowded environments
such as the cytoplasm. This was addressed by the development of a
detailed model for the cytoplasm of the bacterium \textit{Mycoplasma
  genitalium} containing 103 million atoms~\cite{Feig2015}. This
model, as well as a smaller version with 13 million atoms, was
submitted to molecular dynamics simulations~\cite{Yu2016a}. Due to the
size of these systems, the simulation times were very short,
\SI{20}{\nano\second} for the complete model and
\SI{140}{\nano\second} for a smaller version, precluding an accurate
characterization of dynamical effects that can take place in such a
large scale. It was noted already some years ago that even
{microsecond} simulations may be too short to probe the dynamics of
large complexes like viruses~\cite{Larsson2012a} since dynamics is
slower in larger systems in general.

In this work, we report an all-atom model of part of the cytoplasm of
\textit{Escherichia coli} was built to reflect the real biological
system, in particular as regards a realistic composition of this
``soup''. We discuss the difficulties that arise when building such a
system and provide a set of Python scripts and files that anyone can
use to build its own crowded systems with custom parameters and
compounds. Additionally, we performed long molecular dynamics
simulations of our cytoplasm model, reaching \SI{3}{\micro\second} of
total simulation time. To the best of our knowledge, this work is the first to explore the
structural dynamics of an all-atom explicit solvent cytoplasm model 
with microsecond scale molecular dynamics simulations.

@@

We found that completely protonating the phosphate groups of ATP and
FBP is enough to prevent their aggregation with Mg$^{2+}$. While RDFs
show that ATP$^{-3}$ and FBP$^{-4}$ aggregate around Mg$^{2+}$
(Fig.~\ref{fig:avoiding_aggregation}A), this is not observed for
ATPH\textsubscript{3} and
FBPH\textsubscript{4}(Fig.~\ref{fig:avoiding_aggregation}B-C).  
% {\bf
%   WERE ANY INTERMEDIATE PROTONATIONS TESTED? E.G. FBPH2(2-) AND ATPH(2-)?
%   COMPLETELY PROTONATING IS UNPHYSICAL AS WELL AT THE PH STUDIED. FOR E COLI THE PH SHOULDBE ROUGHLY 7.4 https://en.wikipedia.org/wiki/Cytosol} {\color{blue}I didn't try intermediate protonations, but this can be done in about 1 week if necessary. The physiological states of FBP and ATP at pH 7.4 would be the ones we used originally in the soup. So, in this sense, any protonated state could be called unphysical or unrealistic. We discuss the tradeoffs of using this unphysical protonated state in the discussion section}

\begin{figure}[H]
\hspace*{-2cm}\includegraphics[scale=0.5]{rdf_mg.pdf}
\caption{RDFs centered around Mg$^{2+}$ showing the probability of
  finding A) metabolites in the cytoplasm model, B) ATP$^{-3}$ and
  ATPH\textsubscript{3} or C) FBP$^{-4}$ and FBPH\textsubscript{4} in
  small independent systems (please see the methods section for
  details).}
\label{fig:avoiding_aggregation}
\end{figure}

\section*{Discussion}\label{sec:dissc}
The main difference between the behavior of the macromolecules that
compose our model in the crowded system and in the dilute condition is
their translational diffusion coefficient, which drops by around
85\% for the proteins and 93\% for the tRNA
(Fig.~\ref{fig:translational_diffusion}A). The slope of the line
generated by fitting a linear regression model to values of
$D_{cyto}/D_{dil}$ is close to zero (slope $= 0.00028$), showing that
the drop in diffusion coefficient is independent of the size of the
macromolecules. We found that the aggregation of ATP$^{-3}$ and
FBP$^{-4}$ around Mg$^{2+}$, which was used as counter-ion for tRNA,
leads to an exaggerated confinement of the tRNA molecules.  Therefore,
tRNA is considered as an outlier in the duffusion calculations. In
fact, the slope of the linear regression line fitted to
$D_{cyto}/D_{dil}$ is reduced by by 50\% and gets even closer to zero
(slope $= 0.00014$) after removing the diffusion coefficient of tRNA
from the data. These data are in agreement with experimental results
showing that the translational diffusion coefficient of Green
Fluorescent Protein (GFP) in {\em E. coli}, as measured by
fluorescence recovery after photo-bleaching (FRAP), decreased 90\%
when compared to {\em in vitro}
measurements~\cite{Elowitz1999,Konopka2006}.

@@

It has been demonstrated that the force fields that are most widely
used to perform molecular dynamics simulations of macromolecules yield
different results for highly concentrated amino acid
solutions~\cite{andrews2013}. Moreover, it has been reported that
simulations of highly concentrated environment leads to aggregation
when additive force fields are used~\cite{Petrov2014a,Abriata2015a,
  Nawrocki2017a} and several approaches have been proposed to
eliminate this
problem~\cite{Best2014a,Piana2015a,Bashardanesh2018b}. Apart from
protein-water interactions, the London-dispersion coefficients in the
force fields have been investigated as well but their effect on
protein-protein interactions seems to be
minor~\cite{Mohebifar2017a,Walters2018a,Bashardanesh2018b}.  A
comparison at the amino acid level shows that none of the Amber family
of force fields combined with different water model can simultaneously
produce high accuracy Gibbs energies of hydration and amino acid side
chain mobility~\cite{HZhang2018a}. Of the different combinations
tested in that work the Amber99sb-ws~\cite{Best2014a} in conjunction
with the TIP4P/2005 water model~\cite{Abascal2005b} was deemed the
best compromise~\cite{HZhang2018a}. Nevertheless, in a related paper
we find signs that the increased interaction between protein and water
may be slightly too strong as it leads to partial unfolding of a
protein~\cite{Bashardanesh2019a}.

@@

As part of this work we are providing Python scripts with a set of
structure and topology files that any researcher can use to build
their own \textit{E. coli} cytoplasm models. Such models can be used
to investigate research questions about the effects of
crowding on specific systems of interest or about crowding
itself. With the files we are providing it is possible, for example,
to add a probe protein to investigate the effect of crowding on it, to
add new macromolecular or small crowders, evaluate the effects of
 temperatures and change the biomolecular concentration to
increase or decrease the intensity of the crowding effect. We
highlight that, since we have already validated it, researchers can
take advantage from building their own cytoplasm models even if they
do not have access to the infrastructure that is necessary to perform
long molecular dynamics simulations. With modest computatioanal power
it is possible, for example, to build many independent snapshots of
the cytoplasm of \textit{E. coli}. All files are publicly available at
http://github.com/dspoel/soup.

>>>>

\newpage
\begin{abstract}
Biomolecular crowding affects the biophysical and biochemical behavior
of macromolecules when compared to the dilute environment present in
experiments made with isolated proteins. Computational modeling and
simulation are useful tools to study how crowding affects the
structural dynamics and biological properties of macromolecules. As
computational power increased, modeling and simulating large scale
all-atom explicit solvent models of the prokaryote cytoplasm become
possible. In this work, we build an atomistic model of the cytoplasm
of \textit{Escherichia coli} composed of 1.5 million atoms and submit
it to a total of \SI{3}{\micro\second} of molecular dynamics
simulations. The properties of biomolecules under crowding conditions
are compared to those from simulations of the individual compounds
under dilute conditions.  The simulation model is found to be
consistent with experimental data about the diffusion coefficient and
stability of macromolecules under crowded conditions. In order to
stimulate further work we provide a Python script and a set of files
that enables other researchers to build their own \textit{E. coli}
cytoplasm models to address questions related to crowding.
\end{abstract}

@@

The intracellular environment is markedly different from the
\textit{in vitro} model systems from which most of our knowledge about
the biochemical and biophysical behavior of macromolecules has been
derived~\cite{Feig2017a}. The \textit{in vivo} microenvironment is
heterogeneous with high concentrations of many different compounds,
leading to spatial constrains that directly affects biological
activities~\citep{ostrowska2019}. This effect, called macromolecular
crowding, influences the stability of many macromolecules
because the equilibrium is shifted towards a stable native structure
as a result of the entropic component given by the excluded volume
effect~\cite{cheung2005}. At the same time, an effective
destabilization due to the enthalpic component of the Gibbs energy has
been observed, due to replacing interactions with water by weaker
interactions with the
crowders~\cite{Feig2011,miklos2011,Wang2012b}. Additionally, crowding
affects the structure and dynamics of water around proteins, which has
a significant effect on their biological
activities~\cite{Harada2012a,king2013}.
While the structure and dynamics of biomolecules is well characterized
\textit{in vitro}, the understanding of the influences of crowded
environments \textit{in vivo} are still evolving.  Techniques such as
nuclear magnetic resonance~\cite{reckel2007,pielak2008} and
fluorescence spectroscopy~\cite{ignatova2004,xie2008,English2011} are
among the most relevant to the field due to their ability to probe
cells in a non destructive way.

A powerful class of methods is computational modeling and simulation.
As the available computational power increases, models become
increasingly complex and are better representations of reality. The
most simple class of computational models apply Brownian Dynamics
calculations to probe macromolecules immersed in a solution of
crowders that are treated as hard spheres with varying radii to
represent different types of molecules~\cite{Ando2010}. These models
usually use an implicit representation of the
solvent~\cite{Mcguffee2010}. More recently, all-atom explicit solvent
molecular dynamics simulations have been employed on crowded systems
composed of several copies of one or a few
macromolecules~\cite{Wang2017c}. However, these models do not reflect
the true heterogeneity of biologically relevant crowded environments
such as the cytoplasm. This was addressed by the development of a
detailed model for the cytoplasm of the bacterium \textit{Mycoplasma
  genitalium} containing 103 million atoms~\cite{Feig2015}. This
model, as well as a smaller version with 13 million atoms, was
submitted to molecular dynamics simulations~\cite{Yu2016a}. Due to the
size of these systems, the simulation times were very short,
\SI{20}{\nano\second} for the complete model and
\SI{140}{\nano\second} for a smaller version, precluding an accurate
characterization of dynamical effects that can take place in such a
large scale. It was estimated, by extrapolating the advances in
computational power in the last decades to the future, that the
atomistic simulation of an entire \textit{Escherichia coli} bacterium
(about 10$^{11}$ atoms) for \SI{1}{\nano\second} will be possible
around the year 2034~\cite{vanGunsteren2006a}. This prediction may be
too pessimistic, seeing that ``only'' a factor of 50 is lacking
between the simulations by Feig and co-workers~\cite{Feig2015} (from
2015) and the target of a whole bacterium for 1 ns. It was noted
already some years ago, however, that even {microsecond} simulations
may be too short to probe the dynamics of large complexes like
viruses~\cite{Larsson2012a,Larsson2012b} since dynamics is slower in
larger systems in general. Although a meaningful simulation of a
bacterium at the atomistic level does not seem to be within reach in
the foreseeable future, simulations of parts of a cell may still be
useful to adress biological questions.

In this work, we report on an all-atom model of part of the cytoplasm
of \textit{Escherichia coli}, built to reflect the real biological
system, in particular as regards a realistic composition of this
``soup''. We discuss the difficulties that arise when building such a
system and provide a set of Python scripts and files that can be used
to build crowded systems with custom parameters and
compounds. Additionally, we performed long molecular dynamics
simulations of our cytoplasm model, reaching \SI{3}{\micro\second} of
total simulation time. To the best of our knowledge, this work is the
first to explore the structural dynamics of an all-atom explicit
solvent cytoplasm model with microsecond scale molecular dynamics
simulations.

@@

We found that completely protonating the phosphate groups of ATP and
FBP is enough to prevent their aggregation with Mg$^{2+}$. While RDFs
show that ATP$^{3-}$ and FBP$^{4-}$ aggregate around Mg$^{2+}$
(Fig.~\ref{fig:avoiding_aggregation}A), this is not observed for
ATPH\textsubscript{3} and
FBPH\textsubscript{4}(Fig.~\ref{fig:avoiding_aggregation}B-C).  
% {\bf
%   WERE ANY INTERMEDIATE PROTONATIONS TESTED? E.G. FBPH2(2-) AND ATPH(2-)?
%   COMPLETELY PROTONATING IS UNPHYSICAL AS WELL AT THE PH STUDIED. FOR E COLI THE PH SHOULDBE ROUGHLY 7.4 https://en.wikipedia.org/wiki/Cytosol} {\color{blue}I didn't try intermediate protonations, but this can be done in about 1 week if necessary. The physiological states of FBP and ATP at pH 7.4 would be the ones we used originally in the soup. So, in this sense, any protonated state could be called unphysical or unrealistic. We discuss the tradeoffs of using this unphysical protonated state in the discussion section}

\begin{figure}[H]
\hspace*{-2cm}\includegraphics[scale=0.5]{rdf_mg.pdf}
\caption{RDFs centered around Mg$^{2+}$ showing the probability of
  finding A) metabolites in the cytoplasm model, B) ATP$^{3-}$ and
  ATPH\textsubscript{3} or C) FBP$^{4-}$ and FBPH\textsubscript{4} in
  small independent systems (please see the methods section for
  details).}
\label{fig:avoiding_aggregation}
\end{figure}

\section*{Discussion}\label{sec:dissc}
The main difference in behavior of the macromolecules that
compose our model between the crowded system and the dilute condition is
their translational diffusion coefficient, which drops by around
85\% for the proteins and 93\% for the tRNA
(Fig.~\ref{fig:translational_diffusion}A). The slope of the line
generated by fitting a linear regression model to values of
$D_{cyto}/D_{dil}$ is close to zero (slope $= 0.00028$), showing that
the drop in diffusion coefficient is independent of the size of the
macromolecules. We found that the aggregation of ATP$^{3-}$ and
FBP$^{4-}$ around Mg$^{2+}$, which was used as counter-ion for tRNA,
leads to an exaggerated confinement of the tRNA molecules.  Therefore,
tRNA is considered an outlier in the duffusion calculations. In
fact, the slope of the linear regression line fitted to
$D_{cyto}/D_{dil}$ is reduced by by 50\% and  even closer to zero
(slope $= 0.00014$) after removing the diffusion coefficient of tRNA
from the data. These data are in agreement with experimental results
showing that the translational diffusion coefficient of Green
Fluorescent Protein (GFP) in {\em E. coli}, as measured by
fluorescence recovery after photo-bleaching (FRAP), decreased 90\%
when compared to {\em in vitro}
measurements~\cite{Elowitz1999,Konopka2006}.

@@

It has been demonstrated that the force fields that are most widely
used to perform molecular dynamics simulations of macromolecules yield
different results for highly concentrated amino acid
solutions~\cite{andrews2013}. Moreover, it has been reported that
simulations of a highly concentrated environment leads to aggregation
when additive force fields are used~\cite{Petrov2014a,Abriata2015a,
  Nawrocki2017a} and several approaches have been proposed to
eliminate this
problem~\cite{Best2014a,Piana2015a,Bashardanesh2018b}. Apart from
protein-water interactions, the London-dispersion coefficients in the
force fields have been investigated as well but their effect on
protein-protein interactions seems to be
minor~\cite{Mohebifar2017a,Walters2018a,Bashardanesh2018b}.  A
comparison at the amino acid level shows that none of the Amber family
of force fields combined with different water model can simultaneously
produce high accuracy Gibbs energies of hydration and amino acid side
chain mobility~\cite{HZhang2018a}. Of the different combinations
tested in that work the Amber99sb-ws~\cite{Best2014a} in conjunction
with the TIP4P/2005 water model~\cite{Abascal2005b} was deemed the
best compromise~\cite{HZhang2018a}. Nevertheless, in a related paper
we find signs that the increased interaction between protein and water
may be slightly too strong as it leads to partial unfolding of a
protein~\cite{Bashardanesh2019a}.

@@

As part of this work we are providing Python scripts with a set of
structure and topology files that any researcher can use to build
their own \textit{E. coli} cytoplasm models. Such models can be used
to investigate research questions about the effects of crowding on
specific systems of interest or about crowding itself. With the files
we are providing it is possible, for example, to add a probe protein
to investigate the effect of crowding on it, to add new macromolecular
or small crowders, evaluate the effects of temperatures and change the
biomolecular concentration to increase or decrease the intensity of
the crowding effect. We highlight that, since we have already
validated it, researchers can take advantage from building their own
cytoplasm models even if they do not have access to the infrastructure
that is necessary to perform long molecular dynamics simulations. With
modest computatioanal power it is possible, for example, to build many
independent snapshots of the cytoplasm of \textit{E. coli}. All files
are publicly available at http://github.com/dspoel/soup facilitating
contributions to the scripts and data files from other workers in the field.
